---
title: "House Price Prediction & Quality Classification"
output:
  html_document:
    toc: true
    toc_depth: 3
---

## Project Overview

This project analyzes residential housing data to understand the factors that influence property values and to build predictive models for house prices and overall property quality.

The workflow includes data preparation, exploratory data analysis, feature engineering, regression modeling, Random Forest modeling, XGBoost classification, class-imbalance handling, model evaluation, and feature importance analysis.

Two predictive problems are addressed:

1. Predicting house sale prices using regression models.
2. Classifying houses by overall quality using machine learning.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

required_packages <- c(
  "readxl",
  "ggplot2",
  "randomForest",
  "xgboost"
)

missing_packages <- required_packages[
  !required_packages %in% rownames(installed.packages())
]

if (length(missing_packages) > 0) {
  install.packages(missing_packages)
}

library(readxl)
library(ggplot2)
library(randomForest)
library(xgboost)
```

## Load Data

The repository contains separate training and test datasets.

```{r}
train_raw <- read.csv("house_prices_train.csv")
test_raw <- read_excel("house_prices_test.xlsx")

test_raw <- as.data.frame(test_raw)

# Convert numeric Excel columns consistently
test_raw[, sapply(test_raw, is.numeric)] <-
  lapply(test_raw[, sapply(test_raw, is.numeric)], as.numeric)

cat("Training data dimensions:", dim(train_raw), "\n")
cat("Test data dimensions:", dim(test_raw), "\n")
```

## Data Quality Checks

```{r}
cat("Missing values in training data:",
    sum(is.na(train_raw)), "\n")

cat("Missing values in test data:",
    sum(is.na(test_raw)), "\n")

cat("Duplicate rows in training data:",
    sum(duplicated(train_raw)), "\n")

cat("Duplicate rows in test data:",
    sum(duplicated(test_raw)), "\n")
```

## Data Preparation

The same preprocessing steps are applied to both the training and test datasets.

```{r}
prepare_data <- function(df) {

  # Remove YearBuilt because YearRemodAdd captures related information
  if ("YearBuilt" %in% names(df)) {
    df$YearBuilt <- NULL
  }

  # Remodel year categories
  df$YearRemodAdd <- cut(
    df$YearRemodAdd,
    breaks = c(1949, 2000, 2011),
    labels = c(
      "on or before 2000",
      "after 2000"
    ),
    include.lowest = TRUE
  )

  # Total rooms categories
  df$TotRmsAbvGrd <- cut(
    df$TotRmsAbvGrd,
    breaks = c(0, 5, 10, 15),
    labels = c(
      "0-5",
      "5-10",
      "10-15"
    ),
    include.lowest = TRUE
  )

  # Bedroom categories
  df$BedroomAbvGr <- cut(
    df$BedroomAbvGr,
    breaks = c(-1, 5, 15),
    labels = c(
      "less than or equal to 5",
      "More than 5"
    ),
    include.lowest = TRUE
  )

  df
}

train_reg <- prepare_data(train_raw)
test_reg <- prepare_data(test_raw)
```

### Factor Variables

```{r}
factor_cols <- c(
  "YrSold",
  "FullBath",
  "HalfBath",
  "BedroomAbvGr",
  "TotRmsAbvGrd",
  "Fireplaces",
  "YearRemodAdd",
  "OverallQual"
)

for (col in factor_cols) {

  train_levels <- unique(as.character(train_reg[[col]]))

  unknown_test_levels <- setdiff(
    unique(as.character(test_reg[[col]])),
    train_levels
  )

  if (length(unknown_test_levels) > 0) {
    stop(
      paste(
        "Unexpected test level in",
        col,
        ":",
        paste(unknown_test_levels, collapse = ", ")
      )
    )
  }

  train_reg[[col]] <- factor(
    train_reg[[col]],
    levels = train_levels
  )

  test_reg[[col]] <- factor(
    test_reg[[col]],
    levels = train_levels
  )
}

summary(train_reg)
```

## Exploratory Data Analysis

### Distribution of House Prices

```{r}
ggplot(train_reg, aes(x = SalePrice)) +
  geom_histogram(
    bins = 30,
    fill = "steelblue",
    color = "black"
  ) +
  labs(
    title = "Distribution of House Sale Prices",
    x = "Sale Price",
    y = "Frequency"
  ) +
  theme_minimal()
```

### Garage Area and Sale Price

```{r}
ggplot(
  train_reg,
  aes(x = GarageArea, y = SalePrice)
) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Garage Area vs Sale Price",
    x = "Garage Area",
    y = "Sale Price"
  ) +
  theme_minimal()
```

### Finished Basement Area and Sale Price

```{r}
ggplot(
  train_reg,
  aes(x = BsmtFinSF1, y = SalePrice)
) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Finished Basement Area vs Sale Price",
    x = "Finished Basement Area",
    y = "Sale Price"
  ) +
  theme_minimal()
```

### Overall Quality Distribution

```{r}
quality_counts <- as.data.frame(
  table(train_reg$OverallQual)
)

names(quality_counts) <- c(
  "OverallQuality",
  "Count"
)

quality_counts$Percentage <-
  100 * quality_counts$Count /
  sum(quality_counts$Count)

ggplot(
  quality_counts,
  aes(
    x = OverallQuality,
    y = Percentage
  )
) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(
      label = paste0(
        round(Percentage, 1),
        "%"
      )
    ),
    vjust = -0.4
  ) +
  labs(
    title = "Distribution of Overall Quality",
    x = "Overall Quality",
    y = "Percentage"
  ) +
  theme_minimal()
```

## Regression Modeling

Two approaches are evaluated for house-price prediction:

- Multiple Linear Regression
- Random Forest Regression

### Evaluation Function

```{r}
regression_metrics <- function(actual, predicted) {

  mse <- mean(
    (actual - predicted)^2
  )

  rmse <- sqrt(mse)

  r_squared <- cor(
    actual,
    predicted
  )^2

  data.frame(
    MSE = mse,
    RMSE = rmse,
    R_Squared = r_squared
  )
}
```

## Multiple Linear Regression

```{r}
lm_model <- lm(
  SalePrice ~ .,
  data = train_reg
)

lm_predictions <- predict(
  lm_model,
  newdata = test_reg
)

lm_results <- regression_metrics(
  test_reg$SalePrice,
  lm_predictions
)

lm_results
```

## Random Forest Regression

```{r}
set.seed(345)

rf_model <- randomForest(
  SalePrice ~ .,
  data = train_reg,
  ntree = 500,
  importance = TRUE
)

rf_predictions <- predict(
  rf_model,
  newdata = test_reg
)

rf_results <- regression_metrics(
  test_reg$SalePrice,
  rf_predictions
)

rf_results
```

## Tuned Random Forest

A small tuning process is performed over different values of `mtry`.

```{r}
set.seed(345)

candidate_mtry <- c(
  2,
  4,
  6,
  8
)

candidate_mtry <- candidate_mtry[
  candidate_mtry <
    ncol(train_reg)
]

oob_errors <- numeric(
  length(candidate_mtry)
)

rf_models <- vector(
  "list",
  length(candidate_mtry)
)

for (i in seq_along(candidate_mtry)) {

  rf_models[[i]] <- randomForest(
    SalePrice ~ .,
    data = train_reg,
    ntree = 500,
    mtry = candidate_mtry[i],
    importance = TRUE
  )

  oob_errors[i] <-
    tail(
      rf_models[[i]]$mse,
      1
    )
}

best_index <- which.min(
  oob_errors
)

best_mtry <- candidate_mtry[
  best_index
]

cat(
  "Best mtry:",
  best_mtry,
  "\n"
)

rf_tuned_model <-
  rf_models[[best_index]]

rf_tuned_predictions <- predict(
  rf_tuned_model,
  newdata = test_reg
)

rf_tuned_results <- regression_metrics(
  test_reg$SalePrice,
  rf_tuned_predictions
)

rf_tuned_results
```

## Regression Model Comparison

```{r}
regression_comparison <- rbind(

  data.frame(
    Model = "Multiple Linear Regression",
    lm_results
  ),

  data.frame(
    Model = "Random Forest",
    rf_results
  ),

  data.frame(
    Model = "Tuned Random Forest",
    rf_tuned_results
  )
)

regression_comparison
```

## Random Forest Feature Importance

```{r}
rf_importance <- importance(
  rf_tuned_model
)

rf_importance_df <- data.frame(
  Feature = rownames(rf_importance),
  Importance = rf_importance[, 1]
)

rf_importance_df <- rf_importance_df[
  order(
    rf_importance_df$Importance,
    decreasing = TRUE
  ),
]

head(
  rf_importance_df,
  10
)
```

# House Quality Classification

The second modeling problem classifies houses according to overall quality.

Properties with an overall quality score of **7 or higher** are assigned to the positive class.

```{r}
train_cls <- prepare_data(train_raw)
test_cls <- prepare_data(test_raw)

train_cls$OverallQual <- ifelse(
  train_raw$OverallQual >= 7,
  1,
  0
)

test_cls$OverallQual <- ifelse(
  test_raw$OverallQual >= 7,
  1,
  0
)
```

## Prepare Categorical Predictors

```{r}
classification_factor_cols <- c(
  "YrSold",
  "FullBath",
  "HalfBath",
  "BedroomAbvGr",
  "TotRmsAbvGrd",
  "Fireplaces",
  "YearRemodAdd"
)

for (col in classification_factor_cols) {

  all_levels <- unique(
    c(
      as.character(train_cls[[col]]),
      as.character(test_cls[[col]])
    )
  )

  train_cls[[col]] <- factor(
    train_cls[[col]],
    levels = all_levels
  )

  test_cls[[col]] <- factor(
    test_cls[[col]],
    levels = all_levels
  )
}
```

## Class Distribution

```{r}
class_distribution <- as.data.frame(
  table(train_cls$OverallQual)
)

names(class_distribution) <- c(
  "Class",
  "Count"
)

class_distribution$Percentage <-
  100 *
  class_distribution$Count /
  sum(class_distribution$Count)

class_distribution
```

```{r}
ggplot(
  class_distribution,
  aes(
    x = Class,
    y = Percentage
  )
) +
  geom_col(fill = "steelblue") +
  geom_text(
    aes(
      label = paste0(
        round(Percentage, 1),
        "%"
      )
    ),
    vjust = -0.4
  ) +
  labs(
    title = "House Quality Class Distribution",
    x = "Quality Class",
    y = "Percentage"
  ) +
  theme_minimal()
```

## XGBoost Helper Functions

This version uses the native XGBoost interface instead of the older `caret::train(method = "xgbTree")` interface.

```{r}
make_xgb_matrices <- function(
  training_data,
  testing_data
) {

  train_x <- training_data[
    ,
    names(training_data) != "OverallQual"
  ]

  test_x <- testing_data[
    ,
    names(testing_data) != "OverallQual"
  ]

  combined <- rbind(
    train_x,
    test_x
  )

  combined_matrix <- model.matrix(
    ~ . - 1,
    data = combined
  )

  n_train <- nrow(
    train_x
  )

  x_train <- combined_matrix[
    1:n_train,
    ,
    drop = FALSE
  ]

  x_test <- combined_matrix[
    (n_train + 1):
      nrow(combined_matrix),
    ,
    drop = FALSE
  ]

  list(
    train = x_train,
    test = x_test
  )
}
```

### Classification Evaluation

```{r}
classification_metrics <- function(
  actual,
  predicted
) {

  actual <- as.integer(
    actual
  )

  predicted <- as.integer(
    predicted
  )

  TP <- sum(
    predicted == 1 &
    actual == 1
  )

  TN <- sum(
    predicted == 0 &
    actual == 0
  )

  FP <- sum(
    predicted == 1 &
    actual == 0
  )

  FN <- sum(
    predicted == 0 &
    actual == 1
  )

  accuracy <-
    (TP + TN) /
    (TP + TN + FP + FN)

  sensitivity <- ifelse(
    TP + FN == 0,
    NA,
    TP / (TP + FN)
  )

  specificity <- ifelse(
    TN + FP == 0,
    NA,
    TN / (TN + FP)
  )

  precision <- ifelse(
    TP + FP == 0,
    NA,
    TP / (TP + FP)
  )

  f1 <- ifelse(
    is.na(precision) |
      is.na(sensitivity) |
      precision + sensitivity == 0,
    NA,
    2 *
      precision *
      sensitivity /
      (precision + sensitivity)
  )

  data.frame(
    Accuracy = accuracy,
    Sensitivity = sensitivity,
    Specificity = specificity,
    Precision = precision,
    F1 = f1
  )
}
```

## XGBoost Training Function

```{r}
fit_xgb_model <- function(
  training_data,
  testing_data,
  seed = 345
) {

  set.seed(seed)

  matrices <- make_xgb_matrices(
    training_data,
    testing_data
  )

  y_train <- as.numeric(
    training_data$OverallQual
  )

  dtrain <- xgb.DMatrix(
    data = matrices$train,
    label = y_train
  )

  dtest <- xgb.DMatrix(
    data = matrices$test
  )

  params <- list(
    objective = "binary:logistic",
    eval_metric = "logloss",
    max_depth = 6,
    eta = 0.1,
    gamma = 0,
    min_child_weight = 1,
    subsample = 1,
    colsample_bytree = 1
  )

  model <- xgb.train(
    params = params,
    data = dtrain,
    nrounds = 100,
    verbose = 0
  )

  probability <- predict(
    model,
    dtest
  )

  prediction <- ifelse(
    probability >= 0.5,
    1,
    0
  )

  list(
    model = model,
    prediction = prediction,
    probability = probability,
    feature_names = colnames(
      matrices$train
    )
  )
}
```

# XGBoost — No Sampling

```{r}
xgb_no_sampling <- fit_xgb_model(
  train_cls,
  test_cls
)

no_sampling_results <-
  classification_metrics(
    test_cls$OverallQual,
    xgb_no_sampling$prediction
  )

no_sampling_results
```

### Confusion Matrix — No Sampling

```{r}
table(
  Predicted =
    xgb_no_sampling$prediction,
  Actual =
    test_cls$OverallQual
)
```

# Oversampling

The minority class is oversampled to create a more balanced training dataset.

```{r}
oversample_binary <- function(
  df,
  target = "OverallQual",
  seed = 345
) {

  set.seed(seed)

  class0 <- df[
    df[[target]] == 0,
  ]

  class1 <- df[
    df[[target]] == 1,
  ]

  target_size <- max(
    nrow(class0),
    nrow(class1)
  )

  class0_over <- class0[
    sample(
      seq_len(nrow(class0)),
      target_size,
      replace = TRUE
    ),
  ]

  class1_over <- class1[
    sample(
      seq_len(nrow(class1)),
      target_size,
      replace = TRUE
    ),
  ]

  result <- rbind(
    class0_over,
    class1_over
  )

  result[
    sample(seq_len(nrow(result))),
  ]
}

train_over <- oversample_binary(
  train_cls
)

table(
  train_over$OverallQual
)
```

## XGBoost — Oversampling

```{r}
xgb_over <- fit_xgb_model(
  train_over,
  test_cls
)

over_sampling_results <-
  classification_metrics(
    test_cls$OverallQual,
    xgb_over$prediction
  )

over_sampling_results
```

### Confusion Matrix — Oversampling

```{r}
table(
  Predicted =
    xgb_over$prediction,
  Actual =
    test_cls$OverallQual
)
```

# Undersampling

The majority class is undersampled to create a balanced training dataset.

```{r}
undersample_binary <- function(
  df,
  target = "OverallQual",
  seed = 345
) {

  set.seed(seed)

  class0 <- df[
    df[[target]] == 0,
  ]

  class1 <- df[
    df[[target]] == 1,
  ]

  target_size <- min(
    nrow(class0),
    nrow(class1)
  )

  class0_under <- class0[
    sample(
      seq_len(nrow(class0)),
      target_size,
      replace = FALSE
    ),
  ]

  class1_under <- class1[
    sample(
      seq_len(nrow(class1)),
      target_size,
      replace = FALSE
    ),
  ]

  result <- rbind(
    class0_under,
    class1_under
  )

  result[
    sample(seq_len(nrow(result))),
  ]
}

train_under <- undersample_binary(
  train_cls
)

table(
  train_under$OverallQual
)
```

## XGBoost — Undersampling

```{r}
xgb_under <- fit_xgb_model(
  train_under,
  test_cls
)

under_sampling_results <-
  classification_metrics(
    test_cls$OverallQual,
    xgb_under$prediction
  )

under_sampling_results
```

### Confusion Matrix — Undersampling

```{r}
table(
  Predicted =
    xgb_under$prediction,
  Actual =
    test_cls$OverallQual
)
```

# Classification Model Comparison

```{r}
classification_comparison <- rbind(

  data.frame(
    Model = "XGBoost - No Sampling",
    no_sampling_results
  ),

  data.frame(
    Model = "XGBoost - Oversampling",
    over_sampling_results
  ),

  data.frame(
    Model = "XGBoost - Undersampling",
    under_sampling_results
  )
)

classification_comparison
```

# XGBoost Feature Importance

Feature importance is calculated from the undersampled XGBoost model.

```{r}
xgb_importance <- xgb.importance(
  feature_names =
    xgb_under$feature_names,
  model =
    xgb_under$model
)

head(
  xgb_importance,
  10
)
```

```{r}
top_importance <- head(
  xgb_importance,
  10
)

ggplot(
  top_importance,
  aes(
    x = reorder(
      Feature,
      Gain
    ),
    y = Gain
  )
) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title =
      "Top XGBoost Feature Importance",
    x = "Feature",
    y = "Gain"
  ) +
  theme_minimal()
```

# Key Insights

The analysis demonstrates several important points:

- House prices depend on a combination of property characteristics rather than a single factor.
- Lot size, garage area, finished basement area, bathrooms, bedrooms, fireplaces, and overall quality can contribute to property value.
- Regression and ensemble approaches can be compared using MSE, RMSE, and R-squared.
- Class imbalance can materially affect classification performance.
- Oversampling and undersampling provide alternative approaches for improving minority-class prediction.
- Feature importance helps identify the variables contributing most strongly to model predictions.

# Session Information

```{r}
sessionInfo()
```
