---
title: "House Price Prediction & Quality Classification"
output: html_document
---

> Portfolio version of a collaborative analytics project. This analysis covers data preparation, exploratory analysis, regression, Random Forest, XGBoost classification, class-imbalance handling, and model evaluation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```
## Libraries

```{r}
library(readxl)
library(GoodmanKruskal)
library(corrplot)
library(performance)
library(parameters)
library(forcats)
library(car)
library(caret)
library(Metrics)
library(tidyverse)
library(tidymodels)
library(glmnet) 
# library(vip)
library(xgboost)
library(ROSE)
# library(DMwR2)
```

## Load Data

```{r}
data1 = read.csv("house_prices_train.csv")
data2 = read_excel("house_prices_test.xlsx")
data2 <- as.data.frame(data2)
data2[, sapply(data2, is.numeric)] <- lapply(data2[, sapply(data2, is.numeric)], as.integer)
dim(data1)
dim(data2)

```

### Missing Values

```{r}
print(sum(is.na(data1)))
print(sum(is.na(data2)))

```

### Duplicate Rows

```{r}
data1[duplicated(data1) | duplicated(data1, fromLast = TRUE), ]
data2[duplicated(data2) | duplicated(data2, fromLast = TRUE), ]

```

### Remove Unnecessary Variables

```{r}
drop_cols = c("YearBuilt")
data1[drop_cols] = NULL
data2[drop_cols] = NULL
```

### Variable Transformations

```{r}
breaks_1 <- c(1949, 2000, 2011)
data1$YearRemodAdd <- cut(data1$YearRemodAdd, 
                          breaks = breaks_1, labels = c("on or before 2000", "after 2000"))
data2$YearRemodAdd <- cut(data2$YearRemodAdd, 
                          breaks = breaks_1, labels = c("on or before 2000", "after 2000"))

breaks_2 <- c(0, 5, 10, 15)
data1$TotRmsAbvGrd <- cut(data1$TotRmsAbvGrd, 
                          breaks = breaks_2, labels = c("0-5", "5-10", "10-15"))
data2$TotRmsAbvGrd <- cut(data2$TotRmsAbvGrd, 
                          breaks = breaks_2, labels = c("0-5", "5-10", "10-15"))

breaks_3 <- c(-1, 5, 15)
data1$BedroomAbvGr <- cut(data1$BedroomAbvGr, 
                          breaks = breaks_3, labels = c("less than or equal to 5", "More than 5"))
data2$BedroomAbvGr <- cut(data2$BedroomAbvGr, 
                          breaks = breaks_3, labels = c("less than or equal to 5", "More than 5"))

```


### Factor Conversion

```{r}
cols = c("YrSold","FullBath",
         "HalfBath","BedroomAbvGr","TotRmsAbvGrd",
         "Fireplaces","YearRemodAdd","OverallQual")
data1[cols] = lapply(data1[cols], factor)
data2[cols] = lapply(data2[cols], factor)
summary(data1)
summary(data2)
```

## Exploratory Data Analysis

```{r}

# House price #
ggplot(data1, aes(x = SalePrice)) +
  geom_histogram(fill = "blue", color = "black") +
  xlab("Sales Price $") +
  ylab("Frequency") +
  ggtitle("Distribution of Sales Price") +
  scale_x_continuous(labels = scales::comma_format(scale = 1e-3, suffix = "K")) +
  theme_minimal()


# Garage size #
ggplot(data1, aes(x = GarageArea)) +
  geom_histogram(fill = "blue", color = "black") +
  xlab("Garage size in squre feet") +
  ylab("Frequency") +
  ggtitle("Distribution of Garage Size")  +
  theme_minimal()


# Finished suqres size #
ggplot(data1, aes(x = BsmtFinSF1)) +
  geom_histogram(fill = "blue", color = "black") +
  xlab("Finish squre feet size") +
  ylab("Frequency") +
  ggtitle("Distribution of Finish squre feet size")  +
  theme_minimal()


# OverallQual
count_table <- table(data1$OverallQual)
count_df <- as.data.frame(count_table)
colnames(count_df)[colnames(count_df) == "Var1"] <- "Overall_Quality"
count_df$Percentage <- count_df$Freq / sum(count_df$Freq) * 100
ggplot(count_df, aes(x = Overall_Quality, y = Percentage, fill = Overall_Quality)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, color = "black") +
  xlab("Overall Quality Level") +
  ylab("Percentage") +
  ggtitle("Overall Quality") 


# YearRemodAdd
count_table <- table(data1$YearRemodAdd)
count_df <- as.data.frame(count_table)
colnames(count_df)[colnames(count_df) == "Var1"] <- "Year_RemodAdd"
count_df$Percentage <- count_df$Freq / sum(count_df$Freq) * 100
ggplot(count_df, aes(x = Year_RemodAdd, y = Percentage, fill = Year_RemodAdd)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, color = "black") +
  xlab("Modified Year") +
  ylab("Percentage") +
  ggtitle("Modified Year") 


# Full bathrooms
count_table <- table(data1$FullBath)
count_df <- as.data.frame(count_table)
colnames(count_df)[colnames(count_df) == "Var1"] <- "FullBath"
count_df$Percentage <- count_df$Freq / sum(count_df$Freq) * 100
ggplot(count_df, aes(x = FullBath, y = Percentage, fill = FullBath)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, color = "black") +
  xlab("Number of full bathrooms") +
  ylab("Percentage") +
  ggtitle("Number of full bathrooms") 


# LotArea vs FullBath
ggplot(data1, aes(x = FullBath, y = LotArea, fill = FullBath)) +
  geom_boxplot() +
  xlab("Category") +
  ylab("Lot size in square feet") +
  ggtitle("Number of full bathrooms vs Lot size in square feet") +
  theme_minimal() +
  theme(legend.position = "none")

# BsmtFinSF1 vs OverallQual
ggplot(data1, aes(x = OverallQual, y = BsmtFinSF1, fill = OverallQual)) +
  geom_boxplot() +
  xlab("Category") +
  ylab("Finished squre size in feet") +
  ggtitle("Overall Quality vs Finished squre size in feet") +
  theme_minimal() +
  theme(legend.position = "none")

# LotArea vs SalePrice vs Number of full baths
ggplot(data1, aes(x = LotArea, y = SalePrice, color = FullBath)) +
  geom_point() +
  labs(title = "Scatterplot", x = "LotArea", y = "SalePrice") +
  xlab("Lot area size in square feet") +
  ylab("Sales Price $") +
  ggtitle("Sales Price vs Lot area size in square feet vs Number of full baths") +
  scale_y_continuous(labels = scales::comma_format(scale = 1e-3, suffix = "K")) +
  theme_minimal()

# GarageArea vs SalePrice vs Fireplaces
ggplot(data1, aes(x = GarageArea, y = SalePrice, color = Fireplaces)) +
  geom_point() +
  labs(title = "Scatterplot", x = "GarageArea", y = "SalePrice") +
  xlab("Garage size size in square feet") +
  ylab("Sales Price $") +
  ggtitle("Sales Price vs Garagae size in square feet vs Number of fire places") +
  scale_y_continuous(labels = scales::comma_format(scale = 1e-3, suffix = "K")) +
  theme_minimal()

# BsmtFinSF1 vs SalePrice vs HalfBath
ggplot(data1, aes(x = BsmtFinSF1, y = SalePrice, color = HalfBath)) +
  geom_point() +
  labs(title = "Scatterplot", x = "BsmtFinSF1", y = "SalePrice") +
  xlab("Finished squre size in feet") +
  ylab("Sales Price $") +
  ggtitle("Sales Price vs Finished squre size in feet vs Number of Half bathrooms") +
  scale_y_continuous(labels = scales::comma_format(scale = 1e-3, suffix = "K")) +
  theme_minimal()


# SalePrice vs YrSold
ggplot(data1, aes(x = YrSold, y = SalePrice, fill = YrSold)) +
  geom_boxplot() +
  xlab("House Sold Year") +
  ylab("Sales Price $") +
  ggtitle("House Sold Year vs Sales Price") +
  scale_y_continuous(labels = scales::comma_format(scale = 1e-3, suffix = "K")) +
  theme_minimal() +
  theme(legend.position = "none")

```


####  Correlation test
##### Spearman Correlation

```{r}
cor_vars <- subset(data1, select = c("LotArea", "GarageArea", "BsmtFinSF1"))
corrplot(cor(cor_vars, method = "spearman"), method = "color", addCoef.col = "black", tl.col = "black")


```

#####  Goodman Kruskal plot

```{r}
selected_vars <- c("FullBath", "HalfBath", "OverallQual", "BedroomAbvGr", "TotRmsAbvGrd", "Fireplaces", "YrSold")
kruskal_vars <- subset(data1, select = selected_vars)
plot(GKtauDataframe(kruskal_vars))

```

## House Price Regression

###  Create a linear regression model

```{r}
lm_model <- train(SalePrice ~ ., data = data1, method = "lm")


```

#### Model Performances


```{r}
predictions <- predict(lm_model, newdata = data2)
mse <- mean((data2$SalePrice - predictions)^2)
cat("MSE:", mse, "\n")
rmse <- sqrt(mse)
print(paste("Root Mean Squared Error: ", rmse))
r_squared_value <- cor(data2$SalePrice, predictions)^2
cat("R-squared:", r_squared_value, "\n")

```

### Random Forest Model - Without Hyper-parameter tuning


```{r}
set.seed(345)
model_rf <- train(SalePrice ~ ., data = data1, method = "rf")

```

####  Random Forest Model - Model Performances


```{r}
predictions_rf_1 <- predict(model_rf, newdata = data2)
mse_rf_1 <- mean((data2$SalePrice - predictions_rf_1)^2)
print(paste("Mean Squared Error: ", mse_rf_1))
rmse_rf_1 <- sqrt(mse_rf_1)
print(paste("Root Mean Squared Error: ", rmse_rf_1))
r_squared_value_rf_1 <- R2(data2$SalePrice, predictions_rf_1)
cat("R-squared:", r_squared_value_rf_1, "\n")

```

###  Random Forest Model - With Hyper-parameter tuning


```{r}
set.seed(345)
ctrl <- trainControl(method = "cv", number = 7)
grid <- expand.grid(mtry = c(2, 4, 6, 8),
                    min_samples_split = c(2, 5, 10, 15),
                    max_depth = c(5, 10, 15, 20))
model_rf_2 <- train(SalePrice ~ ., data = data1, method = "rf", trControl = ctrl)

```

#### Random Forest Model - Model Performances


```{r}
predictions_rf_2 <- predict(model_rf_2, newdata = data2)
mse_rf_2 <- mean((data2$SalePrice - predictions_rf_2)^2)
rmse_rf_2 <- sqrt(mse_rf_2)
print(paste("Mean Squared Error: ", mse_rf_2))
print(paste("Root Mean Squared Error: ", rmse_rf_2))
r_squared_value_rf_2 <- R2(data2$SalePrice, predictions_rf_2)
cat("R-squared:", r_squared_value_rf_2, "\n")

```


## House Quality Classification


#### Load the data sets

```{r}
data1 = read.csv("data/house_prices_train.csv")
data2 = read_excel("data/house_prices_test.xlsx")
data2 <- as.data.frame(data2)
data2[, sapply(data2, is.numeric)] <- lapply(data2[, sapply(data2, is.numeric)], as.integer)
```


#### Drop columns


```{r}
drop_cols = c("YearBuilt")
data1[drop_cols] = NULL
data2[drop_cols] = NULL
```

#### Variable Conversions

```{r}
breaks_1 <- c(1949, 2000, 2011)
data1$YearRemodAdd <- cut(data1$YearRemodAdd, 
                               breaks = breaks_1, labels = c("on or before 2000", "after 2000"))
data2$YearRemodAdd <- cut(data2$YearRemodAdd, 
                              breaks = breaks_1, labels = c("on or before 2000", "after 2000"))

breaks_2 <- c(0, 5, 10, 15)
data1$TotRmsAbvGrd <- cut(data1$TotRmsAbvGrd, 
                          breaks = breaks_2, labels = c("0-5", "5-10", "10-15"))
data2$TotRmsAbvGrd <- cut(data2$TotRmsAbvGrd, 
                          breaks = breaks_2, labels = c("0-5", "5-10", "10-15"))

breaks_3 <- c(-1, 5, 15)
data1$BedroomAbvGr <- cut(data1$BedroomAbvGr, 
                          breaks = breaks_3, labels = c("less than or equal to", "More than 5"))
data2$BedroomAbvGr <- cut(data2$BedroomAbvGr, 
                          breaks = breaks_3, labels = c("less than or equal to", "More than 5"))

```


#### Factorized

```{r}
cols = c("YrSold","FullBath",
         "HalfBath","BedroomAbvGr","TotRmsAbvGrd",
         "Fireplaces","YearRemodAdd")
data1[cols] = lapply(data1[cols], factor)
data2[cols] = lapply(data2[cols], factor)
```

#### Create the Response and other missing levels

```{r}
training_levels_1 <- levels(data1$FullBath)
training_levels_2 <- levels(data1$BedroomAbvGr)
training_levels_3 <- levels(data1$Fireplaces)
data2$FullBath <- forcats::fct_expand(data2$FullBath, training_levels_1)
data2$BedroomAbvGr <- forcats::fct_expand(data2$BedroomAbvGr, training_levels_2)
data2$Fireplaces <- forcats::fct_expand(data2$Fireplaces, training_levels_3)


data1$OverallQual <- cut(data1$OverallQual,breaks = c(-Inf, 6, Inf),labels = c("0", "1"))
data2$OverallQual <- cut(data2$OverallQual,breaks = c(-Inf, 6, Inf),labels = c("0", "1"))
data1$OverallQual <-factor(data1$OverallQual,levels = c("0","1"),labels = c(0,1))
data2$OverallQual <- factor(data2$OverallQual,levels = c("0","1"),labels = c(0,1))
data1$OverallQual <- relevel(data1$OverallQual, ref = "1")
data2$OverallQual <- relevel(data2$OverallQual, ref = "1")
response_col_name <- "OverallQual"
test_labels <- factor(data2[, response_col_name])
```


#### Imbalanceness of the response variable

```{r} 

count_table <- table(data1$OverallQual)
count_df <- as.data.frame(count_table)
colnames(count_df)[colnames(count_df) == "Var1"] <- "Overall_Quality"
count_df$Percentage <- count_df$Freq / sum(count_df$Freq) * 100
ggplot(count_df, aes(x = Overall_Quality, y = Percentage, fill = Overall_Quality)) +
  geom_bar(stat = "identity", color = "black") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), vjust = -0.5, color = "black") +
  xlab("Overall Quality Level") +
  ylab("Percentage") +
  ggtitle("Overall Quality") 
```

#### XGBoost Model (Without any sampling) - With hyper parameter tuning


```{r}
customSummary <- function(data, lev = NULL, model = NULL) {
  confusionMatrixObj <- confusionMatrix(data$pred, data$obs, positive = lev[1])
  precision <- posPredValue(data$pred, data$obs, positive = lev[1])
  recall <- confusionMatrixObj$byClass["Sensitivity"]
  F1 <- 2 * precision * recall / (precision + recall)
  
  out <- c(precision, recall, F1)
  names(out) <- c("Precision", "Recall", "F1-Score")
  out
}

ctrl <- trainControl(
  method = "cv",
  number = 5,
  search = 'grid',
  savePredictions = "all",
  summaryFunction = customSummary
)

grid <- expand.grid(
  nrounds = c(50, 100),         # Number of boosting rounds
  max_depth = c(3, 6, 9),          # Maximum tree depth
  eta = c(0.01, 0.1, 1),           # Learning rate
  gamma = 0,                    # Minimum loss reduction for further splits
  colsample_bytree = c(0, 1),         # Subsample ratio of columns when constructing each tree
  min_child_weight = c(0, 1),         # Minimum sum of instance weight (hessian) needed in a child
  subsample = c(1, 4, 8)                 # Subsample ratio of the training instance
)

set.seed(345)
model1 <- train(
  OverallQual ~ .,
  data = data1,
  method = "xgbTree", 
  trControl = ctrl,
  tuneGrid = grid
)

```


#### XGBoost Model (Without any sampling) - With hyper parameter tuning (Model Performances)


```{r}
predictions1 <- predict(model1, data2)
confusionMatrix(predictions1, data2$OverallQual, mode = "everything")

```


#### XGBoost Model (Oversampling) - With hyper parameter tuning


```{r}
set.seed(345)
table(data1$OverallQual)
data3 <- ovun.sample(OverallQual ~ ., data = data1, method = "over", N = 1116)$data
set.seed(345)
model2 <- train(
  OverallQual ~ .,
  data = data3,
  method = "xgbTree", 
  trControl = ctrl,
  tuneGrid = grid
)

```


#### XGBoost Model (Oversampling) - With hyper parameter tuning (Model Performances)


```{r}
predictions2 <- predict(model2, data2)
confusionMatrix(predictions2, data2$OverallQual, mode = "everything")

```


#### XGBoost Model (Undersampling) - With hyper parameter tuning 


```{r}
set.seed(345)
data4 <- ovun.sample(OverallQual ~ ., data = data1, method = "under", N = 684)$data
model3 <- train(
  OverallQual ~ .,
  data = data4,
  method = "xgbTree", 
  trControl = ctrl,
  tuneGrid = grid
)

```


#### XGBoost Model (Undersampling) - With hyper parameter tuning (Model Performances)


```{r}
predictions3 <- predict(model3, data2)
confusionMatrix(predictions3, data2$OverallQual, mode = "everything")

```


#### VIP plot XGBoost Model (Undersampling)


```{r}

vip(model3$finalModel, num_features = 10) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 20, face = "bold"),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12)
  )

```
